---
title: "Projekt"
author: "Denisa Neacsu, Binnur Özay, Christofer Richard, Pinar Cavus"
date: "4/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("/Users/hp/Desktop/Skin Cancer/TRA Daten")

TRA_Human_gtex=read.csv(file="tra.2017.human.gtex.5x.table.tsv",sep="\t")
TRA_Mouse_4301=read.csv(file="tra.2014.mouse.4301.5x.table.tsv", sep="\t")
TRA_Human_roth=read.csv(file="tra.2014.human.roth.5x.table.tsv", sep="\t")
TRA_Human_Novartis=read.csv(file="tra.2014.human.5x.table.tsv", sep="\t")
TRA_Mouse_Novartis=read.csv(file="tra.2014.mouse.5x.table.tsv", sep="\t")
TRA_Human_protien_atlas=read.csv(file="Human_protein_atlas_TRA_5median_genes_annotated.tsv", sep=",")
```

```{r}
tissue=TRA_Human_Novartis[,11]
ind=which(tissue=="skin")
TRA.symbol=TRA_Human_Novartis$gene.symbol
TRA.chrx=TRA_Human_Novartis[,11]

skin.TRA1=TRA.symbol[ind]
skin.TRA1
```

```{r}
tissue=TRA_Human_protien_atlas[,11]
ind=which(tissue=="skin")
TRA.symbol=TRA_Human_protien_atlas$Symbol
TRA.chrx=TRA_Human_protien_atlas[,11]

skin.TRA1=TRA.symbol[ind]
skin.TRA1
```

#There is something wrong here that we couldn't figure out; even though there is skin data named "skin" in this table, we still get the output "character(0)" :/

```{r}
tissue=TRA_Human_gtex[,10]
ind=which(tissue=="skin")
TRA.symbol=TRA_Human_gtex$ensembl.symbol
TRA.chrx=TRA_Human_gtex[,10]

skin.TRA1=TRA.symbol[ind]
skin.TRA1
```

#We also get the output "character(0)" here, but that is because this table actually doesn't have any skin data in it, so I guess it's fine.

```{r}
tissue=TRA_Human_roth[,11]
ind=which(tissue=="skin")
TRA.symbol=TRA_Human_roth$gene.symbol
TRA.chrx=TRA_Human_roth[,11]

skin.TRA1=TRA.symbol[ind]
skin.TRA1
```

#For the two mouse datasets we couldn't find any tissue named "skin", but we did find tissue named "epidermis".
#I'm guessing that is because the tables show results of mice and not humans, so the researchers decided to name the sublayer of the skin itself?

```{r}
tissue=TRA_Mouse_4301[,11]
ind=which(tissue=="epidermis")
TRA.symbol=TRA_Mouse_4301$gene.symbol
TRA.chrx=TRA_Mouse_4301[,11]

skin.TRA1=TRA.symbol[ind]
skin.TRA1
```

```{r}
tissue=TRA_Mouse_Novartis[,11]
ind=which(tissue=="epidermis")
TRA.symbol=TRA_Mouse_Novartis$gene.symbol
TRA.chrx=TRA_Mouse_Novartis[,11]

skin.TRA1=TRA.symbol[ind]
skin.TRA1
```

