---
title: "Projekt"
author: "Denisa Neacsu, Binnur Özay, Christofer Richard, Pinar Cavus"
date: "4/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#The first line below is from my computer personally, your home user name and file names are probably different in your computer!

```{r}
setwd("/Users/hp/Desktop/Skin Cancer/TRA Daten")

TRA_Human_gtex=read.csv(file="tra.2017.human.gtex.5x.table.tsv",sep="\t")
TRA_Mouse_4301=read.csv(file="tra.2014.mouse.4301.5x.table.tsv", sep="\t")
TRA_Human_roth=read.csv(file="tra.2014.human.roth.5x.table.tsv", sep="\t")
TRA_Human_Novartis=read.csv(file="tra.2014.human.5x.table.tsv", sep="\t")
TRA_Mouse_Novartis=read.csv(file="tra.2014.mouse.5x.table.tsv", sep="\t")
TRA_Human_protien_atlas=read.csv(file="Human_protein_atlas_TRA_5median_genes_annotated.tsv", sep=",")
```

```{r}
tissue1=TRA_Human_Novartis[,11]
ind1=which(tissue=="skin")
TRA.symbol1=TRA_Human_Novartis$gene.symbol
TRA.chrx1=TRA_Human_Novartis[,11]

skin.TRA1=TRA.symbol[ind1]
skin.TRA1
```

```{r}
tissue2=TRA_Human_protien_atlas[,11]
ind2=which(tissue=="skin")
TRA.symbol2=TRA_Human_protien_atlas$Symbol
TRA.chrx2=TRA_Human_protien_atlas[,11]

skin.TRA2=TRA.symbol[ind2]
skin.TRA2
```

#TRA_Human_gtex for skin sun exposed

```{r}
tissue3.s=TRA_Human_gtex[,10]
ind3.s=which(tissue=="Skin - Sun Exposed")
TRA.symbol3.s=TRA_Human_gtex$ensembl.symbol
TRA.chr3.sx=TRA_Human_gtex[,10]

skin.TRA3.s=TRA.symbol[ind3.s]
skin.TRA3.s
```

#TRA_Human_gtex for skin not sun exposed

```{r}
tissue3.ns=TRA_Human_gtex[,10]
ind3.ns=which(tissue=="Skin - Not Sun Exposed")
TRA.symbol3.ns=TRA_Human_gtex$ensembl.symbol
TRA.chrx3.ns=TRA_Human_gtex[,10]

skin.TRA3.ns=TRA.symbol[ind3.ns]
skin.TRA3.ns
```

#We got the output "character(0)" here, but that is because this table actually doesn't have any skin data in it, so I guess it's fine.

```{r}
tissue4=TRA_Human_roth[,11]
ind4=which(tissue4=="skin")
TRA.symbol4=TRA_Human_roth$gene.symbol
TRA.chrx4=TRA_Human_roth[,11]

skin.TRA4=TRA.symbol[ind4]
skin.TRA4
```

#For the two mouse datasets we couldn't find any tissue named "skin", but we did find tissue named "epidermis".
#I'm guessing that is because the tables show results of mice and not humans, so the researchers decided to name the sublayer of the skin itself?

```{r}
tissue5=TRA_Mouse_4301[,11]
ind5=which(tissue5=="epidermis")
TRA.symbol5=TRA_Mouse_4301$gene.symbol
TRA.chrx5=TRA_Mouse_4301[,11]

skin.TRA5=TRA.symbol5[ind5]
skin.TRA5
```

```{r}
tissue6=TRA_Mouse_Novartis[,11]
ind6=which(tissue6=="epidermis")
TRA.symbol6=TRA_Mouse_Novartis$gene.symbol
TRA.chrx6=TRA_Mouse_Novartis[,11]

skin.TRA6=TRA.symbol6[ind6]
skin.TRA6
```

