---
title: "The Role of Skin-Specific TRAs on Skin Cancer"
author: "Binnur Ozay, Christofer Richard, Denisa Neacsu, Pinar Cavus"
date: "7/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The goal of this project is to find the expression of skin specific genes in skin cancer datasets. As datasets, we used microarray data (Affymetrix chips).

# 1. TRA Analysis
## 1.1. Selection of skin-specific TRAs

Six TRA datasets were used to find skin-specific genes for humans and mice.

```{r}
#Define work directory
projectPath = dirname(rstudioapi::getSourceEditorContext()$path)

#Import TRA Data
setwd(paste(projectPath, 'TRA Daten', sep = '/'))
TRA_Human_gtex = read.csv(file = "tra.2017.human.gtex.5x.table.tsv", sep = "\t")
TRA_Mouse_4301 = read.csv(file = "tra.2014.mouse.4301.5x.table.tsv", sep = "\t")
TRA_Human_roth = read.csv(file = "tra.2014.human.roth.5x.table.tsv", sep = "\t")
TRA_Human_Novartis = read.csv(file = "tra.2014.human.5x.table.tsv", sep = "\t")
TRA_Mouse_Novartis = read.csv(file = "tra.2014.mouse.5x.table.tsv", sep = "\t")
TRA_Human_protien_atlas = read.csv(file = "Human_protein_atlas_TRA_5median_genes_annotated.tsv", sep = ",")

#Skin-specific genes from TRA_Human_Novartis
tissue1 = TRA_Human_Novartis[,11]
ind1 = which(tissue1 == "skin")
TRA.symbol1 = TRA_Human_Novartis$gene.symbol
skin.TRA1 = TRA.symbol1[ind1]

#Skin-specific genes from TRA_Human_protein_atlas
tissue2 = TRA_Human_protien_atlas[,11]
ind2 = which(tissue2 == "skin")
TRA.symbol2 = TRA_Human_protien_atlas$Symbol
skin.TRA2 = TRA.symbol2[ind2]

#Skin-specific genes from TRA_Human_gtex (sun exposed)
tissue3.s = TRA_Human_gtex[,10]
ind3.s = which(tissue3.s == "Skin - Sun Exposed")
TRA.symbol3.s = TRA_Human_gtex$ensembl.symbol
skin.TRA3.s = TRA.symbol3.s[ind3.s]

#Skin-specific genes from TRA_Human_gtex (not sun exposed)
tissue3.ns = TRA_Human_gtex[,10]
ind3.ns = which(tissue3.ns == "Skin - Not Sun Exposed")
TRA.symbol3.ns = TRA_Human_gtex$ensembl.symbol
skin.TRA3.ns = TRA.symbol3.ns[ind3.ns]

#Skin-specific genes from TRA_Human_roth (no skin-specific genes)
tissue4 = TRA_Human_roth[,11]
ind4 = which(tissue4 == "skin")
TRA.symbol4 = TRA_Human_roth$gene.symbol
skin.TRA4 = TRA.symbol4[ind4]

#Skin-specific genes from TRA_Mouse_4301 (epidermis-specific genes)
tissue5 = TRA_Mouse_4301[,11]
ind5 = which(tissue5 == "epidermis")
TRA.symbol5 = TRA_Mouse_4301$gene.symbol
skin.TRA5 = TRA.symbol5[ind5]

#Skin-specific genes from TRA_Mouse_Novartis (epidermis-specific genes)
tissue6 = TRA_Mouse_Novartis[,11]
ind6 = which(tissue6 == "epidermis")
TRA.symbol6 = TRA_Mouse_Novartis$gene.symbol
skin.TRA6 = TRA.symbol6[ind6]
```

## 1.2. Compilation of all skin-specific TRAs

Skin-specific TRAs for humans were put into a single vector and data was cleaned up.

```{r}
#Creating vector with all the human skin genes
skin.genes.human = unique(c(skin.TRA1, skin.TRA2, skin.TRA3.ns, skin.TRA3.s))

#Removing NAs and empty values
skin.genes.human = skin.genes.human[-which(is.na(skin.genes.human))] 
skin.genes.human = skin.genes.human[-which(skin.genes.human == "")] 

#Verification that all NA and empty values were removed
which(is.na(skin.genes.human)) 
which(skin.genes.human == "")

#Saving this vector with all skin-specific human TRAs in a csv file
setwd(paste(projectPath, 'tables', sep = '/'))
write.csv(matrix(skin.genes.human), file = "skin.specific.genes.human.TRA", row.names = F, sep = "\t")
```

# I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```


