---
title: "The Role of Skin-Specific TRAs on Skin Cancer"
author: "Binnur Özay, Christofer Richard, Denisa Neacsu, Pinar Cavus"
date: "Data Analysis Project SS21"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# INTRODUCTION

Skin cancer makes up most of the malignancies all around the world, and consists of two main different types; melanoma and non-melanoma (Linares et al., 2015). Both are mainly caused by exposure to UV lights, but non-melanoma cancer develops in the outer layers of the skin (Leiter & Garbe, 2008), whereas melanoma develops from the melanocytes (Smith et al., 2003). Melanomas only make up a small amount of reported skin cancer diagnoses, however they still account for the majority of skin cancer deaths, making them extremely dangerous (Smith et al., 2003). Melanomas have the highest rate of metastatic capacity of all tumors when they have delayed diagnoses (Potrony et al., 2015). 
More than half of all melanoma cases contain a mutation of the BRAF gene (Alqathama, 2020). The B-Raf protein is a signaling molecule in the Mitogen-activated protein kinase (MAPK) pathway, activating MEK and leading to the phosphorylation of ERK (Alqathama, 2020). This kinase cascade is known to direct important cell processes such as proliferation, differentiation, and survival (Alqathama, 2020).
In this analysis, data from the melanoma cell line WM266-4 is used. This cell line has a BRAF-V600D mutation, meaning that the valine at codon 600 is changed into aspartic acid. This mutation leads to an overexpression of BRAF, therefore an uncontrolled increase in oncogenic signaling (Alqathama, 2020). The cell line is treated with Trametinib, which is an inhibitor of MEK 1 and MEK2 (Zeiser, 2014), ERK1/2-inhibitor, and doxycycline-shERK1 for different periods of time; 3 and 24 hours, 3 and 7 days. Later, the mRNA from these cells are extracted and processed into labeled cRNA fragments. These fragments are hybridized on microarrays and scanned, resulting in expression level quantifications of large amounts of genes. It is to mention that there are no units in this data, because it is simply light intensity scanned. 
The focus in this microarray analysis is a mechanism cancer cells have been shown to use; the upregulation of tissue restricted antigen (TRA) genes. TRAs are highly expressed selectively in their tissues, and T-cells reacting to these self-antigens result in autoimmune diseases (Kyewski & Derbinski, 2004). So that immature T-cells can gauge the self-reactivity of their antigen receptors, they go through a selection in the thymus, where T-cells reacting to TRAs undergo apoptosis (Klein et al., 2014). TRA genes are mostly upregulated in cancer cells for undiscovered reasons, making them good potential drug targets for therapy (Rosenberg, 1999). Current treatments for melanomas include chemotherapy, radiation, immunotherapy, and surgical excision (Gruber & Zito, 2020). Developing a new therapy, where the upregulated skin specific TRA genes are attacked, therefore other tissues don’t get as damaged, would be a significant discovery in the medical field.
Therefore, the main aim of this project is to find upregulated skin specific genes in skin cancer. By looking into the gene expression data of TRA genes in a melanoma cell line, and comparing the variety of the expression levels between different treatments and treatment times, it is possible to gain insight into how the genes affect the survival of the cancer cells. The genes which present significant changes in their expression would be potential drug targets to further investigate, as they would be helping the cancer cell survive under the treatment.


# METHODS AND RESULTS

Quality Control and Normalization of the Data

The programming language R Studio Version 1.4.1106 (R Core Team, 2021) is used throughout this entire project. First, all of the data to be analyzed is downloaded; the numerical microarray data with gene expression intensities of the melanoma cell line and the nominal TRA data (Dinkelacker, 2007;2019) with names of genes originating from different tissues. The TRA data has other information such as chromosome numbers of the given genes which will be put to use later.

```{r}
# 1. TRA Analysis
## 1.1. Selection of skin-specific TRAs

#Define work directory
projectPath = dirname(rstudioapi::getSourceEditorContext()$path)

#Import TRA Data
setwd(paste(projectPath, 'TRA Daten', sep = '/'))
TRA_Human_gtex = read.csv(file = "tra.2017.human.gtex.5x.table.tsv", sep = "\t")
TRA_Mouse_4301 = read.csv(file = "tra.2014.mouse.4301.5x.table.tsv", sep = "\t")
TRA_Human_roth = read.csv(file = "tra.2014.human.roth.5x.table.tsv", sep = "\t")
TRA_Human_Novartis = read.csv(file = "tra.2014.human.5x.table.tsv", sep = "\t")
TRA_Mouse_Novartis = read.csv(file = "tra.2014.mouse.5x.table.tsv", sep = "\t")
TRA_Human_protien_atlas = read.csv(file = "Human_protein_atlas_TRA_5median_genes_annotated.tsv", sep = ",")
```
TRA data is made up of genes from different tissues, different cell lines of humans and mice. All skin genes are extracted and combined in a vector, all NA values are erased. As the skin data available on mice is very little, only the expression of human genes is analyzed. One set of TRA data has the information of sun exposure, as this information might be valuable further into the analysis, these genes are separated into two vectors; exposed to the sun and not exposed to the sun.

```{r}
#Skin-specific genes from TRA_Human_Novartis
tissue1 = TRA_Human_Novartis[,11]
ind1 = which(tissue1 == "skin")
TRA.symbol1 = TRA_Human_Novartis$gene.symbol
skin.TRA1 = TRA.symbol1[ind1]

#Skin-specific genes from TRA_Human_protein_atlas
tissue2 = TRA_Human_protien_atlas[,11]
ind2 = which(tissue2 == "skin")
TRA.symbol2 = TRA_Human_protien_atlas$Symbol
skin.TRA2 = TRA.symbol2[ind2]

#Skin-specific genes from TRA_Human_gtex (sun exposed)
tissue3.s = TRA_Human_gtex[,10]
ind3.s = which(tissue3.s == "Skin - Sun Exposed")
TRA.symbol3.s = TRA_Human_gtex$ensembl.symbol
skin.TRA3.s = TRA.symbol3.s[ind3.s]

#Skin-specific genes from TRA_Human_gtex (not sun exposed)
tissue3.ns = TRA_Human_gtex[,10]
ind3.ns = which(tissue3.ns == "Skin - Not Sun Exposed")
TRA.symbol3.ns = TRA_Human_gtex$ensembl.symbol
skin.TRA3.ns = TRA.symbol3.ns[ind3.ns]

#Skin-specific genes from TRA_Human_roth (no skin-specific genes)
tissue4 = TRA_Human_roth[,11]
ind4 = which(tissue4 == "skin")
TRA.symbol4 = TRA_Human_roth$gene.symbol
skin.TRA4 = TRA.symbol4[ind4]

#Skin-specific genes from TRA_Mouse_4301 (epidermis-specific genes)
tissue5 = TRA_Mouse_4301[,11]
ind5 = which(tissue5 == "epidermis")
TRA.symbol5 = TRA_Mouse_4301$gene.symbol
skin.TRA5 = TRA.symbol5[ind5]

#Skin-specific genes from TRA_Mouse_Novartis (epidermis-specific genes)
tissue6 = TRA_Mouse_Novartis[,11]
ind6 = which(tissue6 == "epidermis")
TRA.symbol6 = TRA_Mouse_Novartis$gene.symbol
skin.TRA6 = TRA.symbol6[ind6]

## 1.2. Compilation of all skin-specific TRAs
#Skin-specific TRAs for humans were put into a single vector and data was cleaned up.
#Creating vector with all the human skin genes

skin.genes.human = unique(c(skin.TRA1, skin.TRA2, skin.TRA3.ns, skin.TRA3.s))

#Removing NAs and empty values
skin.genes.human = skin.genes.human[-which(is.na(skin.genes.human))] 
skin.genes.human = skin.genes.human[-which(skin.genes.human == "")] 

#Verification that all NA and empty values were removed
which(is.na(skin.genes.human)) 
which(skin.genes.human == "")

#Saving this vector with all skin-specific human TRAs in a csv file
setwd(paste(projectPath, 'tables', sep = '/'))
write.csv(matrix(skin.genes.human), file = "skin.specific.genes.human.TRA", row.names = F, sep = "\t")
```
A pie chart is finalized to visualize how much of the TRA data comes from skin. It is found out that skin specific genes make up only 4.84% of all TRA data.
```{r}
#Piechart
#Creating small vectors for each table containing only NON-skin-specific TRAs

#TRA_Human_Novartis
bla1 = which(tissue1 != "skin")
TRA.symbolA = TRA_Human_Novartis$gene.symbol
non.skin.TRA1 = TRA.symbolA[bla1]

#TRA_Human_protein_atlas
bla2 = which(tissue2 != "skin")
TRA.symbolB = TRA_Human_protien_atlas$Symbol
non.skin.TRA2 = TRA.symbolB[bla2]

#TRA_Human_gtex (sun exposed)
bla3.s = which(tissue3.s != "skin")
TRA.symbolC = TRA_Human_gtex$ensembl.symbol
non.skin.TRA3.s = TRA.symbolC[bla3.s]

#TRA_Human_gtex (not sun exposed)
bla3.ns = which(tissue3.ns != "skin")
TRA.symbolD = TRA_Human_gtex$ensembl.symbol
non.skin.TRA3.ns = TRA.symbol3.ns[bla3.ns]

#Putting all the small non-skin-specific TRA vectors in a general vector containing all of them
NON_skin_specific_TRA = unique(c(non.skin.TRA1, non.skin.TRA2, non.skin.TRA3.s, non.skin.TRA3.ns))

#Calculating the total number of TRA
total = length(skin.genes.human) + length(NON_skin_specific_TRA)
prozent = round(length(skin.genes.human)/total*100, 2)
prozent_2 = round(length(NON_skin_specific_TRA)/total*100, 2)

labels = c("Skin-specific TRAs", "Other TRAs")
labels = paste(labels, c(prozent, prozent_2))
labels = paste(labels,"%", sep = "")

#Make pie chart
library(RColorBrewer)
color = brewer.pal(length(labels), "Spectral") 

pie(c(prozent, prozent_2), labels = labels, col = color, main = "Pie Chart skin specific TRAs")
legend("topright", labels, cex = 0.8, fill = color)

#Saving the pie chart
setwd(paste(projectPath, 'plots', sep = '/'))
dev.copy2eps(file="Piechart_skin_TRA.eps") # eps file
dev.copy2pdf(file="Piechart_skin_TRA.pdf") # pdf file
```
Before the analysis, the data must be investigated to see if there are any problems within, for instance a  broken microarray chip, that would alter the results.
```{r, fig.show = 'hide'}
#2. Quality control of microarray data of skin cancer
#The Affymetrix Human Genome U133 Plus 2.0 Array was used and the package for annotation of the chips was downloaded from the Brainarray website.

##2.1. Loading the skin cancer dataset
#15 chips were chosen to be analysed and were downloaded from the GEO website. The 15 chips that were chosen to be used are: GSM1387513, GSM1387514, GSM1387515, GSM1387525, GSM1387527, GSM1387537, GSM1387538, GSM1387544, GSM1387545, GSM1387552, GSM1387553, GSM1387554, GSM1387558, GSM1387559 and GSM1387560.

#Loading the necessary libraries
library(affy)
library(vsn)
library(AnnotationDbi)
library(hgu133plus2hsenstcdf)
library(hgu133plus2hsenstprobe)
library(hexbin)
library(stringr)

#Read all 15 microarray chips
setwd(paste(projectPath, 'rawdata/GSE57721 melanoma', sep = '/'))
data.skin.cancer = ReadAffy()
data.skin.cancer@cdfName = "HGU133Plus2_Hs_ENST"

orig = colnames(exprs(data.skin.cancer))
new.name = substr(orig, 1, nchar(orig)-4)
colnames(exprs(data.skin.cancer)) = new.name

#Saving raw data as a .rda file
setwd(paste(projectPath, 'sessions/rda', sep = '/'))
save.image(file = "data.skin.cancer.rda")
```
Single chips are controlled by visualizing raw microarray data, there aren’t any abnormalities visible in all of the fifteen chips.
```{r,fig.show = 'hide'}
## 2.2. Single chip control
#Visualizing raw microarray data
par(mfrow = c(3,5))
image(data.skin.cancer,col=rainbow(100,start = 0, end = 0.75) [100:1]) #All of the chips seem to have a very good quality, so we can use all of them
```
Gene expression profiling results are highly influenced by RNA quality and degradation (Opitz et al., 2010), therefore RNA degradation plots are made to investigate. All of the lines, each representing a chip, showed similar slope values, and there isn’t any deviation visible. For this reason, all chips are to be included in the analysis as they are all qualitative.
```{r,fig.show = 'hide'}
## 2.2. RNA degradation
# RNA degradation plot shifted and scaled
rna_degradation = AffyRNAdeg(data.skin.cancer)

plotAffyRNAdeg(rna_degradation, col = rainbow(150))
title(sub="Human skin cancer raw")

#Saving shifted and scaled RNA degradation plot
setwd(paste(projectPath, 'plots', sep = '/'))
dev.copy2eps(file = "RNA_degradation_data_skin_cancer_raw.eps")

#RNA degradation plot only shifted
plotAffyRNAdeg(rna_degradation, col = rainbow(150), transform = "shift.only")
title(sub = "Human skin cancer raw")

#Saving only-shifted RNA degradation plot
setwd(paste(projectPath, 'plots', sep = '/'))
dev.copy2eps(file= "RNA_deg_data_skin_cancer_raw.eps")

#Analysis of values and plots from RNA degradation
summaryAffyRNAdeg(rna_degradation) # The values of the slopes are similar, so all of the chips are usable for further analysis

## According to the plots, there is no obvious deviation, no line stands out; all of the chips have high expression intensity. In otherwords, all chips are of high quality.
```
The microarray data is normalized using the package vsn (v3.58.0; Hüber, NA) to remove any systematic variation that can affect the analysis. The pre-normalization and normalized data, which is glog2 transformed, are compared in the forms of boxplots and histograms, in order to see how much systematic variation existed. The boxplots didn’t show a significant difference, even before normalization the median lines are almost in a linear line, indicating that the chips didn’t have much variation to begin with. In both pre- and after normalization histograms, the chip lines follow the same path, and none of the chips stand out.
```{r,fig.show = 'hide'}
## 2.3. Normalization of microarray data
#Data normalization
data.skin.cancer.norm = vsnrma(data.skin.cancer)

#Saving normalized microarray data as a .rda file
setwd(paste(projectPath, 'sessions/rda', sep = '/'))
save.image(file = "normalized_data.rda")
```
```{r,fig.show = 'hide'}
##Boxplots of pre-normalized and normalized data
#Before normalization
par(las = 2)
mmi = c(1.8, 0.7, 1.05, 0.54)
par(mai = mmi)
boxplot(data.skin.cancer, col=rainbow(150), cex.axis = 0.5, main = "Gene expression in skin cancer before normalization")

#Saving boxplot with data before normalization
setwd(paste(projectPath, 'plots', sep = '/'))
dev.copy2eps(file = "boxplot_skin_cancer_rawdata.eps")

#After normalization
par(las = 2, mai = mmi)
boxplot(exprs(data.skin.cancer.norm), col=rainbow(150), cex.axis = 0.5, main = "Gene expression in skin cancer after normalization")

#Saving boxplot with data after normalization
setwd(paste(projectPath, 'plots', sep = '/'))
dev.copy2eps(file = "boxplot_skin_cancer_normalized.eps")
```
```{r,fig.show = 'hide'}
##Histogram
#Histogram before data normalization
hist(data.skin.cancer, col = rainbow(150), main = "Histogram data pre-normalized", xlab = "Log intensity", ylab = "Density")

#Saving histogram before normalization
setwd(paste(projectPath, 'plots', sep = '/'))
dev.copy2eps(file = "Histogram data pre-normalized")

#Histogram after data normalization
for (i in 1:ncol(exprs(data.skin.cancer.norm))) if (i == 1) plot(density(exprs(data.skin.cancer.norm)[, i]), col = "red", xlab = "Log intensity", ylab = "Density") else lines(density(exprs(data.skin.cancer.norm)[, i]), col = "red")

#Saving histogram after normalization
setwd(paste(projectPath, 'plots', sep = '/'))
dev.copy2eps(file = "Histogram data normalized.eps")
```
After the normalization, a meanSdPlot is made, and from this plot, it is seen that the mean values of the chips are almost linear, once again proving that they are high quality chips.
```{r,fig.show = 'hide'}
## 2.4. meanSd Plot
#Plotting meanSdPlot
meanSdPlot(data.skin.cancer.norm)

#Saving meanSd plot
setwd(paste(projectPath, 'plots', sep = '/'))
dev.copy2eps(file = "meanSdPlot_data_skin_cancer_normalized.eps")
```
Cleaning Data and Creating a Dataframe

In the data from microarray chips genes are saved as numbers, to be able to view and analyse the data easier, these numbers are switched with their names using the ensembl IDs that were downloaded from https://www.ensembl.org/biomart . Later, all expression values are extracted from the microarray data, and put into a data frame with chip IDs as columns and gene symbols as rows.
```{r,fig.show = 'hide'}
# 3. Analysis
## 3.1. Creating the gene expression data frame

#Extraction of the expression values
Expression = exprs(data.skin.cancer.norm)

#Select only ENST IDs, which represent the transcript IDs
enst_id = str_detect(rownames(Expression),"ENST")
Expression = Expression[enst_id,]

#Remove .x_at suffix from each row name
rownames(Expression) = sub("\\..*", "", rownames(Expression)) 

#Remove .CEL suffix from each column name
colnames(Expression) = sub("\\..*", "", colnames(Expression))

#Load ensemble IDs and gene symbols from annotation data
setwd(paste(projectPath, 'tables', sep = '/'))
Ensembl = read.csv(paste(projectPath, 'tables/ensembl.103.txt', sep = '/'), sep = "\t")

#Extracting the transcript IDs and the HGNC symbols (gene symbols)
transcriptIDs = Ensembl$Transcript.stable.ID
HGNC_symbol = Ensembl$HGNC.symbol

#Assigning every gene symbol to a transcript ID
names(HGNC_symbol) = transcriptIDs

#Replace ensemble IDs with HGNC symbol
chipIDs = rownames(Expression)
newChipIDs = chipIDs[chipIDs %in% transcriptIDs]
Expression = Expression[newChipIDs,]

symbol = HGNC_symbol[newChipIDs]
rownames(Expression) = as.character(symbol)
head(Expression)

#Cleaning the Expression data frame (removing NAs and empty values "")
length(which(is.na(Expression)))
length(which(rownames(Expression) == ""))

all.spaces.exp = sapply(rownames(Expression), function(x){which(x == "")})
Expression = Expression[-which(all.spaces.exp>0),] 
rm(all.spaces.exp)

```
After empty spaces and NA values are removed from the data, skin specific antigens are extracted to create the main data frame of skin TRA expression levels in skin cancer to analyze. The column names are changed so that the treatment and time point of every chip is easily viewed.
```{r}
## 3.2. Extraction of skin-specific TRAs from the gene expression data frame
#Filtering the gene expression data frame to obtain expression values only for skin-specific TRAs
skin.genes.human.expression = which(rownames(Expression)%in%skin.genes.human)
data.skin.cancer.human_skin_TRA = Expression[skin.genes.human.expression,]

y = sort(rownames(data.skin.cancer.human_skin_TRA))
data.skin.cancer.human_skin_TRA = data.skin.cancer.human_skin_TRA[unique(y[y !=""]),]

#Renaming the columns in the data frame with the skin-specific genes
colnames(data.skin.cancer.human_skin_TRA)[1] = "Sherk1_D3a"
colnames(data.skin.cancer.human_skin_TRA)[2] = "Sherk1_D3b"
colnames(data.skin.cancer.human_skin_TRA)[3] = "Sherk1_D3c"
colnames(data.skin.cancer.human_skin_TRA)[4] = "Sherk1_D7a"
colnames(data.skin.cancer.human_skin_TRA)[5] = "Sherk1_D7c"
colnames(data.skin.cancer.human_skin_TRA)[6] = "ERK12_inh_3ha"
colnames(data.skin.cancer.human_skin_TRA)[7] = "ERK12_inh_3hb"
colnames(data.skin.cancer.human_skin_TRA)[8] = "Trametinib_3hb"
colnames(data.skin.cancer.human_skin_TRA)[9] = "Trametinib_3hc"
colnames(data.skin.cancer.human_skin_TRA)[10] = "ERK12_inh_24ha"
colnames(data.skin.cancer.human_skin_TRA)[11] = "ERK12_inh_24hb"
colnames(data.skin.cancer.human_skin_TRA)[12] = "ERK12_inh_24hc"
colnames(data.skin.cancer.human_skin_TRA)[13] = "Trametinib_24ha"
colnames(data.skin.cancer.human_skin_TRA)[14] = "Trametinib_24hb"
colnames(data.skin.cancer.human_skin_TRA)[15] = "Trametinib_24hc"

head(data.skin.cancer.human_skin_TRA)
```
Before getting to the gene expression analysis a bar plot is created to find out which chromosome the skin specific TRAs are mostly located at. Most skin related genes are usually located on chromosome 1 (Volz et al., 1993), which also seems to be the case in this data.
```{r,fig.show = 'hide'}
## 3.4. Barplot
#Barplot
#Get all the chromosome numbers for all human skin TRAs
chromosome_1 = TRA_Human_Novartis[,c(3,7)]
chromosome_skin1 = as.matrix(chromosome_1[ind1,])

chromosome_2 = TRA_Human_protien_atlas[,c(6,2)]
chromosome_skin2 = as.matrix(chromosome_2[ind2,])

chromosome_3 = TRA_Human_gtex[,c(3,6)]
chromosome_skin3 = as.matrix(chromosome_3[c(ind3.s, ind3.ns),])

#Combine all 3 tables and remove duplicates and empty values to make a clean list of chromosome numbers
chromosome_gene_table = rbind(chromosome_skin1, chromosome_skin2, chromosome_skin3)
rownames(chromosome_gene_table) = c()
a = sort(chromosome_gene_table[,1])

rownames(chromosome_gene_table) = chromosome_gene_table[,1]
chromosome_gene_table = chromosome_gene_table[,2]
chromosome_list = chromosome_gene_table[unique(a[a !=""])]

#Count the number of genes on each chromosome
localization = table(chromosome_list)
localization = as.matrix(localization)
sorted_chromosomes = str_sort(rownames(localization), numeric = T)
chromosome_counts = localization[sorted_chromosomes,]

#Make barplot
chromosome = c(1:22, "X")

barplot(chromosome_counts, main = "Distribution of skin-specific TRAs over chromosomes", xlab = "Chromosomes", ylab = "Number of skin-specific TRAs", names.arg = chromosome, cex.names = 0.7, col = "darkred")

#Saving the barplot
setwd(paste(projectPath, 'plots', sep = '/'))
dev.copy2eps(file = 'Distribution of skin-specific TRAs over chromosomes.eps') # eps file
dev.copy2pdf(file = 'Distribution of skin-specific TRAs over chromosomes.pdf') # pdf file
```
A heatmap was made to visualize the expression data of all skin-specific TRAs over all chips.
```{r,fig.show = 'hide'}
## 3.3. Heatmap
#Heatmap
library(gplots)

heatmap.2(data.skin.cancer.human_skin_TRA, key = T, col = redgreen(100), symkey = F, symm = F,symbreaks = T, scale = "row", density.info = "none", trace = "none", cexRow = 0.6, cexCol = 0.6)

#Saving the heatmap
setwd(paste(projectPath, 'plots', sep = '/'))
dev.copy2eps(file = 'Heatmap for skin specific TRAs in skin cancer_1.eps') #eps file
dev.copy2pdf(file = 'Heatmap for skin specific TRAs in skin cancer_1.pdf') #pdf file
```
To begin the analysis the gene expression is visualized with box plots. There are too many genes overall, therefore the boxes look like overlapping lines, which does not give much information. Instead, only the highly expressed genes are chosen to be visualized. A highly expressed gene doesn’t have to be an upregulated gene, and an upregulated gene has the possibility of not being a highly expressed gene. A threshold of twelve is set according to the overall expression levels to define the high expression margin. Aside from the two highest expressed genes GPNMB and ANXA2, the genes NDRG1, RPL18, DCT, CD44, and MLANA show varying expression levels between chips with levels above the threshold in some chips and below in others.
```{r,fig.show = 'hide'}
#Expression of all skin-specific TRAs in skin cancer
par(las=2)
boxplot(t(data.skin.cancer.human_skin_TRA), col = rainbow(length(rownames(data.skin.cancer.human_skin_TRA))), main = "Expression of all skin genes in melanoma", cex.axis = 0.8)
```
```{r}
#Putting the highly expressed genes in a vector
highgenes = c()
for (i in length(colnames(data.skin.cancer.human_skin_TRA))) {
  highgenes = c(highgenes, names(which(data.skin.cancer.human_skin_TRA[,i] > 12)))
}
highgenes = unique(highgenes)

high = data.skin.cancer.human_skin_TRA[highgenes,]

# Expression of the highly expressed genes
par(las=2)
boxplot(t(high),col=rainbow(length(rownames(high))),main="Highly expressed skin genes in melanoma",cex.axis=0.8) #It can be clearly observed that there are two genes with distinctively highest expressions: GPNMB and ANXA2.
```
Upregulated Genes
In order to determine the mostly upregulated genes, the expression data has to be transformed by using a log fold change. Downregulated genes was also determined, as interesting results may arise after comparison. To further confirm whether the genes are significantly up- or downregulated, Welsh t-test was applied as a statistical test. Here, the null hypothesis states that the genes are not differently expressed between the two time points after treatment.There isn’t any data available from cells without any treatment, therefore for chips that have been treated with ERK1/2-inhibitor or Trametinib, the gene expression between the three hour and twenty four hour time points is compared to see which genes showed a higher expression after twenty one hours of treatment. The zero point for dox-shERK1 chips is taken as three days, as there isn’t any earlier data available.
```{r}
## 3.7. Log fold change and significance
### 3.7.1. Upregulated genes
#LogFC: Finding the most upregulated genes from each treatment
#Average the gene expression for each treatment at each time point
sherk_3d = apply(data.skin.cancer.human_skin_TRA[,1:3], 1, mean)
sherk_7d = apply(data.skin.cancer.human_skin_TRA[,4:5], 1, mean)
erk12_3h = apply(data.skin.cancer.human_skin_TRA[,6:7], 1, mean)
erk12_24h = apply(data.skin.cancer.human_skin_TRA[,10:12], 1, mean)
trametinib_3h = apply(data.skin.cancer.human_skin_TRA[,8:9], 1, mean)
trametinib_24h = apply(data.skin.cancer.human_skin_TRA[,13:15], 1, mean)

#Calculate the log fold change for each gene from each treatment
logFC_sherk = sort(sherk_7d - sherk_3d, decreasing = T)
logFC_erk12 = sort(erk12_24h - erk12_3h, decreasing = T)
logFC_trametinib =  sort(trametinib_24h - trametinib_3h, decreasing = T)

#Top logFC from each treatment
logFC_sherk_top = logFC_sherk[1:11]
logFC_erk12_top = logFC_erk12[1:10]
logFC_trametinib_top = logFC_trametinib[1:10]
```
Not all genes encountered at the end of this analysis are significantly upregulated to become drug targets. Therefore, ten genes with the highest log-ratios from different treatments are chosen to be filtered through a Welsh t-test. The following upregulated genes are respectively from dox-shERK1, ERK1/2 inhibitor, and Trametinib chips. 
```{r}
#T-test: Filtering upregulated genes that are significant
#Checking significance for genes in the top logFC of Dox-shERK
t_sherk = function(x) {
  t.test(data.skin.cancer.human_skin_TRA[x, 1:3], data.skin.cancer.human_skin_TRA[x, 4:5])$p.value
}
t_test_sherk = sort(sapply(names(logFC_sherk), t_sherk), decreasing = F)
t_test_sherk_significant = t_test_sherk[which(t_test_sherk < 0.05)]

upreg_sherk_significant = names(logFC_sherk_top)[names(logFC_sherk_top) %in% names(t_test_sherk_significant)]
upreg_sherk_significant

#Checking significance for genes in the top logFC of ERK1/2 inhibitor
t_erk12 = function(x) {
  t.test(data.skin.cancer.human_skin_TRA[x, 6:7], data.skin.cancer.human_skin_TRA[x, 10:12])$p.value
}
t_test_erk12 = sort(sapply(names(logFC_erk12), t_erk12), decreasing = F)
t_test_erk12_significant = t_test_erk12[which(t_test_erk12 < 0.05)]

upreg_erk12_significant = names(logFC_erk12_top)[names(logFC_erk12_top) %in% names(t_test_erk12_significant)]
upreg_erk12_significant

#Checking significance for genes in the top logFC of trametinib
t_trametinib = function(x) {
  t.test(data.skin.cancer.human_skin_TRA[x, 8:9], data.skin.cancer.human_skin_TRA[x, 13:15])$p.value
}
t_test_trametinib = sort(sapply(names(logFC_trametinib), t_trametinib), decreasing = F)
t_test_trametinib_significant = t_test_trametinib[which(t_test_trametinib < 0.05)]

upreg_trametinib_significant = names(logFC_trametinib_top)[names(logFC_trametinib_top) %in% names(t_test_trametinib_significant)]
upreg_trametinib_significant
```
Box plots were built to visualize the expressions of the significantly upregulated genes.
```{r,fig.show = 'hide'}
##Boxplots for upregulated genes
#Dox-shERK
data_sherk_upreg = data.skin.cancer.human_skin_TRA[upreg_sherk_significant, c(1:5)]
data.new_sherk = rbind(data_sherk_upreg, data_sherk_upreg)
data.new_sherk[c(1:2), c(4:5)] = NA
data.new_sherk[c(3:4), c(1:3)] = NA
order.vector = sort(colnames(t(data.new_sherk)),index.return=T)$ix

boxplot(t(data.new_sherk)[,order.vector], cex.axis = 0.6, col = c("blue", "pink"), main = "Upregulated genes in treatment with Dox-shERK1", ylab = "log2(gene expression)", xlab = "Upregulated genes", las = 2)
abline(h = c(6:12), col = "grey", lty = 3)
legend("topleft", c("Dox-shERK1 after 3 days", "Dox-shERK1 after 7 days"), col = c("blue", "pink"), pch = 15, bg = "white")
```
```{r,fig.show = 'hide'}
#ERK1/2 inhibitor
data_erk12_upreg = data.skin.cancer.human_skin_TRA[upreg_erk12_significant, c(6:7, 10:12)]
data.new_erk12 = rbind(data_erk12_upreg, data_erk12_upreg)
data.new_erk12[c(1:8), c(3:5)] = NA
data.new_erk12[c(9:16), c(1:2)] = NA
order.vector = sort(colnames(t(data.new_erk12)),index.return=T)$ix

boxplot(t(data.new_erk12)[,order.vector], cex.axis = 0.6, col = c("blue", "pink"), main = "Upregulated genes in treatment with ERK1/2 inhibitor", ylab = "log2(gene expression)", xlab = "Upregulated genes", las = 2)
abline(h = c(6:14), col = "grey", lty = 3)
legend("topleft",c("ERK1/2 inhibitor after 3h", "ERK1/2 inhibitor after 24h"), col = c("blue", "pink"), pch = 15, bg = "white")
```
```{r}
#Trametinib
data_trametinib_upreg = data.skin.cancer.human_skin_TRA[upreg_trametinib_significant, c(8:9, 13:15)]
data.new_trametinib = rbind(data_trametinib_upreg, data_trametinib_upreg)
data.new_trametinib[c(1:8), c(3:5)] = NA
data.new_trametinib[c(9:16), c(1:2)] = NA
order.vector = sort(colnames(t(data.new_trametinib)),index.return=T)$ix

boxplot(t(data.new_trametinib)[,order.vector], cex.axis = 0.6, col = c("blue", "pink"), main = "Upregulated genes in treatment with Trametinib", ylab = "log2(gene expression)", xlab = "Upregulated genes", las = 2)
abline(h = c(6:14), col = "grey", lty = 3)
legend("left",c("Trametinib after 3h", "Trametinib after 24h"),col=c("blue", "pink"), pch = 15, bg = "white")
```
The package “Limma” (v4.1, ???) is used as a model of comparison to the upregulated genes that were found manually using the log fold change. Limma takes the normalized data and performs statistical tests to discover any quantitative changes in the expression (Ritchie et al., 2015). The results show mostly identical genes to the already found upregulated genes. 
```{r,fig.show = 'hide'}
## Limma Analysis
library(limma)

#First group: We compare the expression values from day 3 to day 7 with the Dox-shERK treatment
# Create design matrix
design = model.matrix(~ 0+factor(c(2,2,2,1,1)))
colnames(design) = c("Sherk7D", "Sherk3D")

#Fit linear model
fit = lmFit(data.skin.cancer.human_skin_TRA[,1:5], design)
contrast.matrix = makeContrasts(Sherk7D-Sherk3D, levels=design)
fit2 = contrasts.fit(fit, contrast.matrix)
fit2 =  eBayes(fit2, 0.01)

#Create top table with the DEGs sorted by log fold change
T1 = topTable(fit2,adjust.method = "fdr", sort.by = "logFC")

#Select only the first 10 genes with a positive log fold change from the table
lola1=T1[which(T1$logFC>0),]
lola1 = lola1[1:10,]

#Confirmed top upregulated genes
intersect(upreg_sherk_significant, rownames(lola1))

#Second group: We compare 3h ERK1/2 and 24h ERK1/2 treatment
#Create design matrix
design_ERK = model.matrix(~ 0+factor(c(1,1,1,2,2)))
colnames(design_ERK) = c("ERK12_inh_24h", "ERK12_inh_3h")

#Fit linear model
data_ERK = data.skin.cancer.human_skin_TRA[,c(10:12, 6,7)]
fit_ERK = lmFit(data_ERK, design_ERK) 
contrast.matrix_ERK = makeContrasts(ERK12_inh_24h-ERK12_inh_3h, levels=design_ERK)
fit2_ERK = contrasts.fit(fit_ERK, contrast.matrix_ERK)
fit2_ERK =  eBayes(fit2_ERK, 0.01)

#Create top table with the DEGs sorted by log fold change
T2 = topTable(fit2_ERK,adjust.method = "fdr", sort.by = "logFC", number = 130) 

#Select only the first 10 genes with a positive log fold change from the table
lola2 = T2[which(T2$logFC>0),]
lola2 = lola2[1:10,]

#Confirmed top upregulated genes
intersect(upreg_erk12_significant, rownames(lola2))

#Third group: We compare 3h to 24h Trametinib treatment
#Create design matrix
design_Trametinib = model.matrix(~ 0+factor(c(1,1,1,2,2)))
colnames(design_Trametinib) = c("Trametinib_24h", "Trametinib_3h")

#Fit linear model
data_Trametinib = data.skin.cancer.human_skin_TRA[,c(13:15, 8,9)]
fit_Trametinib = lmFit(data_Trametinib, design_Trametinib) 
contrast.matrix_Trametinib = makeContrasts(Trametinib_24h-Trametinib_3h, levels=design_Trametinib)
fit2_Trametinib= contrasts.fit(fit_Trametinib, contrast.matrix_Trametinib)
fit2_Trametinib =  eBayes(fit2_Trametinib, 0.01)

#Create top table with the DEGs sorted by log fold change
T3 = topTable(fit2_Trametinib,adjust.method = "fdr", sort.by = "logFC", number = Inf ) 

#Select only the first 10 genes with a positive log fold change from the table
lola3 = T3[which(T3$logFC>0),]
lola3 = lola3[1:10,]

#Confirmed top upregulated genes
intersect(upreg_trametinib_significant, rownames(lola3))


#The genes from limma analysis were very similar to the ones found manually, which means that the manual way succeeded. This was checked by doing the intersection of the vectors for the significant upregulated genes of each treatment and the top 10 genes found with limma.
```
The existence of significantly downregulated genes indicates that the treatment has been successful, and that these genes were inducing tumors before the treatment. That is why, ten genes from different treatments with the lowest log-ratio are filtered out and evaluated under t-tests, and visualized via box plots. The following downregulated genes are respectively from dox-shERK1, ERK1/2 inhibitor, and Trametinib chips. 
```{r}
### 3.7.2. Downregulated genes
#Least logFC from each treatment
logFC_sherk_down_regulated = logFC_sherk[503:512]
logFC_erk12_down_regulated = logFC_erk12[503:512]
logFC_trametinib_down_regulated = logFC_trametinib[503:512]

#T-test: Filtering downregulated genes that are significant
#Checking significance for genes in the last 10 logFC of Dox-shERK
t_sherk = function(x) {
  t.test(data.skin.cancer.human_skin_TRA[x, 1:3], data.skin.cancer.human_skin_TRA[x, 4:5])$p.value
}
t_test_sherk = sort(sapply(names(logFC_sherk), t_sherk), decreasing = F)
t_test_sherk_significant = t_test_sherk[which(t_test_sherk < 0.05)]

downreg_sherk_significant = names(logFC_sherk_down_regulated)[names(logFC_sherk_down_regulated) %in% names(t_test_sherk_significant)]
downreg_sherk_significant

#Checking significance for genes in the last 10 logFC of ERK1/2 inhibitor
t_erk12 = function(x) {
  t.test(data.skin.cancer.human_skin_TRA[x, 6:7], data.skin.cancer.human_skin_TRA[x, 10:12])$p.value
}
t_test_erk12 = sort(sapply(names(logFC_erk12), t_erk12), decreasing = F)
t_test_erk12_significant = t_test_erk12[which(t_test_erk12 < 0.05)]

downreg_erk12_significant = names(logFC_erk12_down_regulated)[names(logFC_erk12_down_regulated) %in% names(t_test_erk12_significant)]
downreg_erk12_significant

## Checking significance for genes in the last 10 logFC of trametinib
t_trametinib = function(x) {
  t.test(data.skin.cancer.human_skin_TRA[x, 8:9], data.skin.cancer.human_skin_TRA[x, 13:15])$p.value
}
t_test_trametinib = sort(sapply(names(logFC_trametinib), t_trametinib), decreasing = F)
t_test_trametinib_significant = t_test_trametinib[which(t_test_trametinib < 0.05)]

downreg_trametinib_significant = names(logFC_trametinib_down_regulated)[names(logFC_trametinib_down_regulated) %in% names(t_test_trametinib_significant)]
downreg_trametinib_significant
```
```{r}
##Boxplots for downregulated genes
#Dox-shERK
data_sherk_downreg = data.skin.cancer.human_skin_TRA[downreg_sherk_significant, c(1:5)]
data.new_sherk_downreg = rbind(data_sherk_downreg, data_sherk_downreg)
data.new_sherk_downreg[c(1:2), c(4:5)] = NA
data.new_sherk_downreg[c(3:4), c(1:3)] = NA
order.vector = sort(colnames(t(data.new_sherk_downreg)),index.return=T)$ix

boxplot(t(data.new_sherk_downreg)[,order.vector], cex.axis = 0.6, col = c("blue", "pink"), main = "Downregulated genes in treatment with Dox-shERK", ylab = "log2(gene expression)", xlab = "Downregulated genes", las = 2)
abline(h = c(6:12), col = "grey", lty = 3)
legend("topleft", c("Dox-shERK after 3 days", "Dox-shERK after 7 days"), col = c("blue", "pink"), pch = 15, bg = "white")
```
```{r,fig.show = 'hide'}
#ERK1/2 inhibitor
data_erk12_downreg = data.skin.cancer.human_skin_TRA[downreg_erk12_significant, c(6:7, 10:12)]
data.new_erk12_downreg = rbind(data_erk12_downreg, data_erk12_downreg)
data.new_erk12_downreg[c(1:8), c(3:5)] = NA
data.new_erk12_downreg[c(9:16), c(1:2)] = NA
order.vector = sort(colnames(t(data.new_erk12_downreg)),index.return=T)$ix

boxplot(t(data.new_erk12_downreg)[,order.vector], cex.axis = 0.6, col = c("blue", "pink"), main = "Downregulated genes in treatment with ERK1/2 inhibitor", ylab = "log2(gene expression)", xlab = "Downregulated genes", las = 2)
abline(h = c(6:14), col = "grey", lty = 3)
legend("topleft",c("ERK1/2 inhibitor after 3h", "ERK1/2 inhibitor after 24h"), col = c("blue", "pink"), pch = 15, bg = "white")
```
```{r,fig.show = 'hide'}
#Trametinib
data_trametinib_downreg = data.skin.cancer.human_skin_TRA[downreg_trametinib_significant, c(8:9, 13:15)]
data.new_trametinib_downreg = rbind(data_trametinib_downreg, data_trametinib_downreg)
data.new_trametinib_downreg[c(1:8), c(3:5)] = NA
data.new_trametinib_downreg[c(9:16), c(1:2)] = NA
order.vector = sort(colnames(t(data.new_trametinib_downreg)),index.return=T)$ix

boxplot(t(data.new_trametinib_downreg)[,order.vector], cex.axis = 0.6, col = c("blue", "pink"), main = "Downregulated genes in treatment with Trametinib", ylab = "log2(gene expression)", xlab = "Downregulated genes", las = 2)
abline(h = c(6:14), col = "grey", lty = 3)
legend("topleft",c("Trametinib after 3h", "Trametinib after 24h"),col=c("blue", "pink"), pch = 15, bg = "white")
```
It should be noted that TYRP1 is significantly upregulated under Trametinib treatment, but significantly downregulated under dox-shERK1.

Genes that are upregullated in skin cancer and are at the same time TRAs are especially interesting, as they pose as potential drug targets. To visualize these upregulated TRAs, a venn diagram was made.
```{r}
## 3.9. Venn diagram
#Finding upregulated genes in skin cancer by logFC
#Average the gene expression for each treatment at each time point
sherk_3d_all = apply(Expression[,1:3], 1, mean)
sherk_7d_all = apply(Expression[,4:5], 1, mean)
erk12_3h_all = apply(Expression[,6:7], 1, mean)
erk12_24h_all = apply(Expression[,10:12], 1, mean)
trametinib_3h_all = apply(Expression[,8:9], 1, mean)
trametinib_24h_all = apply(Expression[,13:15], 1, mean)

## Calculate the log fold change for each gene from each treatment
logFC_sherk_all = sherk_7d_all - sherk_3d_all
logFC_erk12_all = erk12_24h_all - erk12_3h_all
logFC_trametinib_all =  trametinib_24h_all - trametinib_3h_all

## Only positive logFC from each treatment (only upregulated genes!)
logFC_sherk_pos = which(logFC_sherk_all > 0)
logFC_erk12_pos = which(logFC_erk12_all > 0)
logFC_trametinib_pos = which(logFC_trametinib_all > 0)

upreg_all = unique(c(names(logFC_sherk_pos), names(logFC_erk12_pos), names(logFC_trametinib_pos)))

# T-test to filter significantly upregulated genes
## Checking significance for Dox-shERK
t_sherk_all = function(x) {
  if (Expression[x,1] == Expression[x,2] | Expression[x,1] == Expression[x,3] | Expression[x,3] == Expression[x,2] | Expression[x,4] == Expression[x,5])
    return(0.5)
  else
  t.test(Expression[x, 1:3], Expression[x, 4:5])$p.value
}
t_test_sherk_all = sapply(names(logFC_sherk_all), t_sherk_all)
t_test_sherk_significant_all = t_test_sherk_all[which(t_test_sherk_all < 0.05)]

upreg_sherk_significant_all = names(logFC_sherk_pos)[names(logFC_sherk_pos) %in% names(t_test_sherk_significant_all)]

## Checking significance for ERK1/2 inhibitor
t_erk12_all = function(x) {
  t.test(Expression[x, 6:7], Expression[x, 10:12])$p.value
}
t_test_erk12_all = sapply(names(logFC_erk12_all), t_erk12_all)
t_test_erk12_significant_all = t_test_erk12_all[which(t_test_erk12_all < 0.05)]

upreg_erk12_significant_all = names(logFC_erk12_pos)[names(logFC_erk12_pos) %in% names(t_test_erk12_significant_all)]

## Checking significance for Trametinib
t_trametinib_all = function(x) {
  t.test(Expression[x, 8:9], Expression[x, 13:15])$p.value
}
t_test_trametinib_all = sapply(names(logFC_trametinib_all), t_trametinib_all)
t_test_trametinib_significant_all = t_test_trametinib_all[which(t_test_trametinib_all < 0.05)]

upreg_trametinib_significant_all = names(logFC_trametinib_pos)[names(logFC_trametinib_pos) %in% names(t_test_trametinib_significant_all)]

#Venn diagram (Upregulated genes that are also TRAs)
library(venn)
list_venn = list(A = unique(c(upreg_sherk_significant_all, upreg_erk12_significant_all, upreg_trametinib_significant_all)), B = skin.genes.human)
names(list_venn) = c("Upregulated genes" , "TRA")

venn(list_venn, box = F, ilab = T, zcolor = c("#0073C2FF", "#EFC000FF"), cex = 1000)
text(500, 900, "Potential drug targets", cex = 2)

#Save the venn diagramm in a pdf file
setwd(paste(projectPath, 'plots', sep = '/'))
pdf("Venn_diagramm_trial.pdf")
venn(list_venn, box = F, ilab = T, zcolor = c("#0073C2FF", "#EFC000FF"))
text(500, 900, "Potential drug targets", cex = 2)
dev.off()

#Upregulated non-skin TRAs in skin cancer
upreg_tra = intersect(unique(c(upreg_sherk_significant_all, upreg_erk12_significant_all, upreg_trametinib_significant_all)), skin.genes.human)
upreg_tra_nonskin = upreg_tra[!upreg_tra%in%rownames(data.skin.cancer.human_skin_TRA)]
length(upreg_tra_nonskin) #There are no non-skin specific TRAs that are upregulated in skin cancer
```
A Wilcoxon signed rank test was run for each chip and a significant difference between the expression levels of the two gene groups was observed. Genes expressed both under the sun and without the sun are removed from the data, and the genes that were only expressed under sun exposure are acquired. Four of these sun exposure related genes are also upregulated genes: ETV7, TGFBI, DAPL1, and ARHGEF5.
```{r,fig.show = 'hide'}
## 3.10. Sun exposure
#Finding the sun exposed and not sun exposed genes in our skin-specific expression dataset
sun = intersect(rownames(data.skin.cancer.human_skin_TRA), skin.TRA3.s)
no.sun = intersect(rownames(data.skin.cancer.human_skin_TRA), skin.TRA3.ns)

#Removing the gene names to make a box plot of each chip
sun_exp = data.skin.cancer.human_skin_TRA[sun,]
nosun_exp = data.skin.cancer.human_skin_TRA[no.sun,]
sunexpchip = unname(sun_exp[,13])
nosunexpchip = unname(nosun_exp[,13])
lmts = range(sunexpchip,nosunexpchip)

#Comparing their significance with a welsh t-test for each chip
ttest_sun = function(x) {
  t.test(data.skin.cancer.human_skin_TRA[sun, x], data.skin.cancer.human_skin_TRA[no.sun, x])$p.value
}
sapply(1:15, ttest_sun)
## All of them have a p value of smaller than 0.05, proving a significant difference between the expression of sun or no sun exposure genes. So now we will compare the expression of genes that are expressed with or without sun exposure.

# Genes that are expressed with or without sun exposure
both.sun = intersect(sun, no.sun)
# Genes that are only sun exposure related
only_sun = sun[!sun%in%both.sun]
#Genes that are only no sun exposure related
only_nosun = no.sun[!no.sun%in%both.sun]

#The upregulated sun exposed genes ETV7 and TGFBI are especially interesting, as ETV7 downregulation is related to poor prognosis (so working against the cancer) and TGFBI is strongly associated with tumor proliferation.

# The visualization of these genes can be seen in the graphs showing the logFC for all significantly upregulated genes.
significant_sun_1 = upreg_sherk_significant[upreg_sherk_significant%in%only_sun]
significant_sun_2 = upreg_erk12_significant[upreg_erk12_significant%in%only_sun]
significant_sun_3 = upreg_trametinib_significant[upreg_trametinib_significant%in%only_sun]

upregulated_sun = unique(c(significant_sun_1, significant_sun_2, significant_sun_3))

significant_nosun_1 = upreg_sherk_significant[upreg_sherk_significant%in%only_nosun]
significant_nosun_2 = upreg_erk12_significant[upreg_erk12_significant%in%only_nosun]
significant_nosun_3 = upreg_trametinib_significant[upreg_trametinib_significant%in%only_nosun]

upregulated_nosun = unique(c(significant_nosun_1, significant_nosun_2, significant_nosun_3))
```
The expression difference of genes exposed to sun and genes that were not exposed to sun is visualized by k-means. After running the elbow, silhouette, and one other method provided by the package NbClust (v3.0; Charrad, 2015), it is decided that two clusters are the optimal amount for this data. After clustering and plotting with the package factoextra (v1.0.3; Kassambara, 2020), two clusters are defined with an average silhouette width of 0.65, indicating that there is indeed a clusterable difference between the expressions of these two gene groups.
```{r,fig.show = 'hide'}
## 4. Clustering
#Load libraries needed for clustering
library(factoextra)
library(clustertend)
library(fpc)
library(cluster)
library(clValid)
library(NbClust)
library(dplyr)

#Deciding on the optimal number of clusters
set.seed(123)
both = data.skin.cancer.human_skin_TRA[c(only_sun, only_nosun),]

#Elbow Method
fviz_nbclust(both, kmeans, method = "wss")+labs(subtitle = "Elbow method")

#Silhouette Method
fviz_nbclust(both, kmeans, method = "silhouette")+labs(subtitle = "Silhouette method")

#NbClust
nb.both = NbClust(both, distance = "euclidean", min.nc = 2,max.nc = 10, method = "kmeans")
fviz_nbclust(nb.both)
#The optimal number of clusters is 2.

#Is the data really clusterable?
hopkins(both, n = nrow(both)-1) #H value is lower than 0.5 and is relatively close to 1, meaning that there is a high probability that there are clusters in the sun/no sun data.

#Cluster Validation: Silhouette Information
km.both = eclust(both, "kmeans", k = 2, nstart = 30, graph = FALSE)
fviz_silhouette(km.both, palette = "jco",
ggtheme = theme_classic())
#Average silhouette width of both clusters are higher than 0.5, meaning they are significantly clustered.

#Visualize k-means clusters
fviz_cluster(km.both, data = both, palette = c("orange", "blue"), geom = "point", ellipse.type = "convex", ggtheme = theme_bw())
```
Principal component analysis (PCA) is carried out on the 24h and 7d chips to find the genes that explain most of the variance in the data and to reduce the amount of dimensions. PC1 and PC2 represent 88% of all the variance of the melanoma dataset. Five genes with the top variance from each of these components are chosen to continue the analysis with: ANXA2, DCT, NDRG1, TYRP1, PTPRZ1 from PC1, and DSP, GJA1, PPP1R14C, TGFBI, UCN2 from PC2.
```{r,fig.show = 'hide'}
## 5. Principal component analysis
#Centering and scaling, pick data only from final time and only genes with highest variance
pca_dataframe = scale(data.skin.cancer.human_skin_TRA)
pca_dataframe = t(pca_dataframe[,c(4:5, 10:15)])

topVar = apply(pca_dataframe, 2, var)
q75 = quantile(topVar, probs = 0.75)
i.topvar = which(topVar >= q75)
pca_dataframe_topVar = pca_dataframe[,i.topvar]

#PCA
skin_pca_topVar = prcomp(pca_dataframe_topVar, scale. = F, center = F)

#Biplot of first 2 principal components
biplot(skin_pca_topVar)

#Proportion of variance of each principal component
variance = (skin_pca_topVar$sdev)^2
prop.variance = variance/sum(variance)
names(prop.variance) = 1:length(prop.variance)
barplot(prop.variance, ylab='Proportion of variance', xlab = "Principal components", main = "Proportion of variance of each principal component")

#Barplots of principal components
par(las = 2)
barplot(skin_pca_topVar$rotation[,1], main = "PC1", ylab = "Contribution", xlab = "Genes")
barplot(skin_pca_topVar$rotation[,2], main = "PC2", ylab = "Contribution", xlab = "Genes")

#Genes with most contribution to PC1 and PC2
top_genes_pca1 = names(sort(abs(skin_pca_topVar$rotation[,1]), decreasing = T)[1:5])
top_genes_pca2 = names(sort(abs(skin_pca_topVar$rotation[,2]), decreasing = T)[1:5])

#Scatterplot of data with PC1 and PC2 as new coordinates
plot(skin_pca_topVar$x[,1], skin_pca_topVar$x[,2], 
     col= c("red", "red", "green", "green", "green", "blue", "blue", "blue"), pch=19,
     xlab='PC1', ylab='PC2', main = "PC1 and PC2 as new coordinates")
legend("topright",c("Sherk", "ERK1/2 inhibitor", "Trametinib"),col=c("red", "green", "blue"), pch = 15, bg = "yellow")
```

## 6. Regression models

Finally, regression models were made to predict the expression of the upregulated TRAs by using the genes previously selected by PCA. It is hoped that the expression of these essential upregulated genes in cancer can be modulated by targeting the genes from PCA. This would then be a good opportunity in developing new drugs with new drug targets.

```{r}

```

## I

You

```{r}

```

## I

You

```{r}

```

DISCUSSION

Regardless of treatment type and time variation between different chips, GPNMB and ANXA2 proved to be highest expressed in all of them. ANXA2 is known as a biomarker in cancers, it is already seen as an ideal target for cancer therapy because of its crucial functions in cancer cell proliferation, migration, invasion, metastasis (Wang et al., 2019). GPNMB has been formerly identified as an aggressive pro-metastatic protein that is highly expressed in many tumor types, including melanoma (Taya & Hammes, 2018). The fact that brings these two together is that they are both responsible for the genesis of melanosomes (Setaluri & Jayanthy, 2013; Delevoye et al., 2015).
Melanoma develops from melanocytes, which normally assemble melanosomes; organelles in mammals responsible for melanin production (Setaluri & Jayanthy, 2013). It makes sense that these two genes related to the assembly and maturation of melanosomes are found to be highly expressed in melanoma. The two other important genes TYRP1 and DCT found in this analysis, which are also involved in the development of melanosomes (Setaluri & Jayanthy, 2013), are both upregulated and principal component genes at the same time. DCT also known as TYRP2 encodes proteins that play roles in the melanin synthesis, just as TYRP1 (Setaluri & Jayanthy, 2013). The upregulation and high expression of these melanosome related genes, indicates that a therapy focusing to direct melanosome functions would be effective against melanoma.
Microphthalmia-associated transcription factor (MITF) is a key signal regulator with the subset target genes GPNMB and TYRP1 (Rose et al., 2016). MITF plays important roles in melanosomal function, pigmentation, but it is also required for sustaining the BRAF-mutant melanoma, and supporting the melanoma cell proliferation (Rose et al., 2016). It has been observed before that MITF related genes are upregulated under certain BRAF and MEK inhibitors, such as Trametinib (Rose et al., 2016). This would explain the expression results of TYRP1 in this project. It is upregulated as it is needed for the maintenance of the melanoma cell, and the Trametinib treatment, which is a MEK1/2 inhibitor further accelerates its expression. The other two treatments are inhibitors of ERK, which partially successfully seem to have worked, because one of the genes supporting the cancer cell is downregulated under them.
Other significant genes that have been found are the four upregulated genes that are only expressed under the sun. DAPL1 is a gene upregulated by the aforementioned MITF (Ma et al., 2018), and TGFBI regulates cyclin-dependent kinases, which are known to be in charge of the cell cycle, and allows a high metastatic potential to the melanoma (Lauden et al., 2014). Lastly, ARHGEF5 codes for a guanin exchange factor that is crucial for certain membrane formations that are related to the invasive potential of cancer cells (Kuroiwa et al., 2011). The knowledge that such genes, aiding the development of melanoma, have only been upregulated under sun exposure, confirms the already commonly known fact that the exposure to sun is a trigger for skin cancer.
The fourth upregulated gene under sun exposure, ETV7, codes for a transcription factor, which is usually seen to be downregulated in melanoma patients and correlates with a poor prognosis. (Qu et al., 2020). However, ETV7 is seen to be upregulated in our data under Trametinib, suggesting that it might have a role in attacking the tumor microenvironment, leading to a better prognosis. The fact that this gene is expressed under the sun suggests that it might be a natural response to the tumorigenic genes, upregulated under sun exposure. Trametinib helps upregulate it further to fight tumor proliferation. The relationship between ETV7 and the three genes mentioned above should be looked into with more detail, as a treatment upregulating its expression seems to be possibly effective against melanoma.
What’s important to give attention to here is that almost all of the upregulated genes are either regulated by MITF or are from the data that was exposed to the sun. This indicates that the high expression of the genes that are related to MITF are as effective as exposure to sun when it comes to helping create or supporting the melanoma. Since all of the genes analysed on are human skin specific TRA genes, a new treatment where the expression of the genes mentioned above are targeted, would not harm healthy tissues. This opens countless opportunities for future research and treatment of melanomas, especially because melanoma is highly metastatic.
Through the opposite expression of TYRP1, and varying up- and downregulated genes between chips of different treatments, the different effects of MEK and ERK inhibitors on melanoma have been shown. Even though these treatments in this project have shown to downregulate some known tumor inducing genes such as WEE1, an important regulator in the cell cycle (Tashnizi et al., 2016), it should be further researched and discussed on if these inhibitors should be used on melanoma. If yes, a combination of other drugs are needed as treatment to inhibit the accelerating effects of MAPK inhibitors on melanosome related genes.
The strong linear dependency found between the expression of the gene DSP and most of the upregulated genes also raises new questions requiring additional research.  DSP codes for a desmosomal protein, which helps human melanoma metastases to escape the immune system (Salerno et al., 2016). Some desmosomal proteins are used to identify melanomas, even when there are not any immune signatures seen (Salerno et al., 2016). This linear relationship between DSP and the upregulated genes could be used for better diagnosis of melanomas, however more information on the expression of DSP in melanomas is needed. 
The findings that melanosome related genes, and genes highly expressed under sun exposure strongly influence the melanoma cells, and that there is a strong linear relationship between the upregulated and the high variance genes set base for further research in new treatments for melanoma. Therefore, it can be stated that this analysis has given successful results.

LITERATURE

Alboukadel, K. (2017). Practical Guide to Cluster Analysis in R: Unsupervised Machine Learning, STHDA
Alqathama A. (2020). BRAF in malignant melanoma progression and metastasis: potentials and challenges. American journal of cancer research, 10(4), 1103–1114.
Delevoye, C., Heiligenstein, X., Ripoll, L., Gilles-Marsens, F., Dennis, M. K., Linares, R. A., Derman, L., Gokhale, A., Morel, E., Faundez, V., Marks, M.S., Raposo, G. (2015). BLOC-1 Brings Together the Actin and Microtubule Cytoskeletons to Generate Recycling Endosomes, Current Biology, Vol 26(1), 1-13, https://doi.org/10.1016/j.cub.2015.11.020
Dinkelacker, 2007. A database of genes that are expressed in a tissue-restricted manner to analyse promiscuous gene expression in medullary thymic epithelial cells. Diplomarbeit, Albert-Ludwigs-Universitaet, Freiburg, Germany.
Dinkelacker, 2019. Chromosomal clustering of tissue restricted antigens, Dissertation, University Heidelberg, Germany.
Gruber, P., Zito, P. M. (2021). Jan, Skin Cancer. [Updated 2020 Nov 19]. In: StatPearls [Internet]. Treasure Island (FL): StatPearls Publishing; Available from: https://www.ncbi.nlm.nih.gov/books/NBK441949/
Klein, L., Kyewski, B., Allen, P. et al. (2014). Positive and negative selection of the T cell repertoire: what thymocytes see (and don't see). Nat Rev Immunol 14, 377–391. https://doi.org/10.1038/nri3667
Kuroiwa, M., Oneyama, C., Nada, S., & Okada, M. (2011). The guanine nucleotide exchange factor Arhgef5 plays crucial roles in Src-induced podosome formation. Journal of cell science, 124(Pt 10), 1726–1738. https://doi.org/10.1242/jcs.080291
Kyewski, B., Derbinski, J. (2004) Self-representation in the thymus: an extended view. Nat Rev Immunol 4, 688–698. https://doi.org/10.1038/nri1436
Lauden, L., Siewiera, J., Boukouaci, W., Ramgolam, K., Mourah, S., Lebbe, C., Charron, D., Aoudjit, F., Jabrane-Ferrat, N., Al-Daccak, R. (2014). TGF-β-Induced (TGFBI) Protein in Melanoma: A Signature of High Metastatic Potential, Journal of Investigative Dermatology, Vol:134(6), 1675-1685, ISSN 0022-202X, https://doi.org/10.1038/jid.2014.20.
Leiter, U., & Garbe, C. (2008). Epidemiology of melanoma and nonmelanoma skin cancer--the role of sunlight. Advances in experimental medicine and biology, 624, 89–103. https://doi.org/10.1007/978-0-387-77574-6_8
Linares, M. A., Zakaria, A., & Nizran, P. (2015). Skin Cancer. Primary care, 42(4), 645–659. https://doi.org/10.1016/j.pop.2015.07.006
Ma, X., Hua, J., Zheng, G., Li, F., Rao, C., Li, H., Wang, J., Pan, L., & Hou, L. (2018). Regulation of cell proliferation in the retinal pigment epithelium: Differential regulation of the death-associated protein like-1 DAPL1 by alternative MITF splice forms. Pigment cell & melanoma research, 31(3), 411–422. https://doi.org/10.1111/pcmr.12676
Moloney, E. B., Moskites, A., Ferrari, E. J., Isacson, O., & Hallett, P. J. (2018). The glycoprotein GPNMB is selectively elevated in the substantia nigra of Parkinson's disease patients and increases after lysosomal stress. Neurobiology of disease, 120, 1–11. https://doi.org/10.1016/j.nbd.2018.08.013
Opitz, L., Salinas-Riester, G., Grade, M., Jung, K., Jo, P., Emons, G., Ghadimi, B. M., Beissbarth, T., & Gaedcke, J. (2010). Impact of RNA degradation on gene expression profiling. BMC medical genomics, 3, 36. https://doi.org/10.1186/1755-8794-3-36
Potrony, M., Badenas, C., Aguilera, P., Puig-Butille, J. A., Carrera, C., Malvehy, J., Puig, S. (2015). Update in genetic susceptibility in melanoma. Annals of translational medicine, 3(15), 210.	 https://doi.org/10.3978/j.issn.2305-5839.2015.08.11
Qu, H., Zhao, H., Zhang, X., Liu, Y., Li, F., Sun, L., & Song, Z. (2020). Integrated Analysis of the ETS Family in Melanoma Reveals a Regulatory Role of ETV7 in the Immune Microenvironment. Frontiers in immunology, 11, 612784. https://doi.org/10.3389/fimmu.2020.612784
Ritchie, M. E., Phipson, B., Wu, D., Hu, Y., Law, C. W., Shi, W., & Smyth, G. K. (2015). limma powers differential expression analyses for RNA-sequencing and microarray studies. Nucleic acids research, 43(7), e47. https://doi.org/10.1093/nar/gkv007
Rose, A. A., Annis, M. G., Frederick, D. T., Biondini, M., Dong, Z., Kwong, L., Chin, L., Keler, T., Hawthorne, T., Watson, I. R., Flaherty, K. T., Siegel, P. M. (2016). MAPK Pathway Inhibitors Sensitize BRAF-Mutant Melanoma to an Antibody-Drug Conjugate Targeting GPNMB, DOI: 10.1158/1078-0432.CCR-16-1192 
Rosenberg S. A. (1999). A new era for cancer immunotherapy based on the genes that encode cancer antigens. Immunity, 10(3), 281–287. https://doi.org/10.1016/s1074-7613(00)80028-x
Salerno, E. P., Bedognetti, D., Mauldin, I. S., Deacon, D. H., Shea, S. M., Pinczewski, J., Obeid, J. M., Coukos, G., Wang, E., Gajewski, T. F., Marincola, F. M., & Slingluff, C. L., Jr (2016). Human melanomas and ovarian cancers overexpressing mechanical barrier molecule genes lack immune signatures and have increased patient mortality risk. Oncoimmunology, 5(12), e1240857. https://doi.org/10.1080/2162402X.2016.1240857
Setaluri, V., Jayanthy, A. (2013). Coat Color Mutations, Animals, Brenner's Encyclopedia of Genetics, Second Edition, Academic Press, 58-60, ISBN 9780080961569, https://doi.org/10.1016/B978-0-12-374984-0.00275-8.
Smith, R. A., Mettlin, C. J., Eyre, H. (2003). Melanoma and Nonmelanoma Skin Cancer, Holland-Frei Cancer Medicine. 6th edition. Hamilton (ON): BC Decker;  https://www.ncbi.nlm.nih.gov/books/NBK13764/
Tashnizi, A. H., Jaberipour, M., Razmkhah, M., Rahnama, S., & Habibagahi, M. (2016). Tumour suppressive effects of WEE1 gene silencing in neuroblastomas. Journal of cancer research and therapeutics, 12(1), 221–227. https://doi.org/10.4103/0973-1482.165861
Taya, M., Hammes, S. R. (2018). Glycoprotein Non-Metastatic Melanoma Protein B (GPNMB) and Cancer: A Novel Potential Therapeutic Target. Steroids, 133, 102–107. https://doi.org/10.1016/j.steroids.2017.10.013
Volz, A., Korge, B. P., Compton, J. G., Ziegler, A., Steinert, P. M., & Mischke, D. (1993). Physical mapping of a functional cluster of epidermal differentiation genes on chromosome 1q21. Genomics, 18(1), 92–99. https://doi.org/10.1006/geno.1993.1430
Wang, T., Wang, Z., Niu, R., & Wang, L. (2019). Crucial role of Anxa2 in cancer progression: highlights on its novel regulatory mechanism. Cancer biology & medicine, 16(4), 671–687. https://doi.org/10.20892/j.issn.2095-3941.2019.0228
Zeiser R. (2014). Trametinib. Recent results in cancer research. Fortschritte der Krebsforschung. Progrès dans les recherches sur le cancer, 201, 241–248. https://doi.org/10.1007/978-3-642-54490-3_15

